\pkg{SVMBridge}: Using SVM Solvers from R

### 2016-02-11 paper/article.tex
small fixes, add sketch of paper temporarily

Support Vector Machines (SVM) are linear classifiers
that try to maxime the margin of a binary problem \citep{cortes1995support}.
As SVMs are only linear classifiers, a kernelized version
is used for more complex data. Basically, they work by
solving the  following convex problem:
\begin{equation}
        \min_{w \in \mathcal{H}, b \in \R} \quad \frac{1}{2} ||w||^2 + C \cdot \sum_{i=1}^n \max \Big( 0, 1 - y_i \big( \langle w, \varphi(x_i) \rangle_{\mathcal{H}} + b \big) \Big).
        \label{SVMproblemPrimal}
\end{equation}
Here 
$\mathcal{D}=\{(x_1, y_1), \dots, (x_n, y_n)\} \in (X \times \{ \pm 1\})^n$
is the labeled data,
and  $\varphi : X \to \mathcal{H}$ is a feature map into a reproducing
kernel Hilbert space $\mathcal{H}$, corresponding to a positive definite
(Mercer) kernel function $k : X \times X \to \R$, fulfilling
$\langle \varphi(x), \varphi(x') \rangle_{\mathcal{H}} = k(x, x')$ for
all $x, x' \in X$. Furthermore, $C > 0$ is a regularization parameter.
It controls the complexity of the SVM model.

Often, the problem is not directly solved, but dualized via Lagrangian theory.
Then the problem can be stated as 
\begin{align}
        \max_{\alpha \in \R^n} \quad & \sum_{i=1}^n \alpha_i - \frac{1}{2} \sum_{i,j=1}^n \alpha_i \alpha_j y_i y_j k(x_i, x_j) \label{SVMproblemDual} \\
        \text{s.t.} \quad & \sum_{i=1}^{n} y_i \alpha_i = 0 \text{~~and~~} 0 \leq \alpha_i \leq C \,\,\, \forall \, i \in \{1, \dots, n\}. \notag
        \enspace
\end{align}
The solution then takes the form $w = \sum_{i=1}^n \alpha_i y_i \varphi(x_i)$
and the offset~$b$ can be computed from the Karush-Kuhn-Tucker (KKT)
complementarity conditions.

SVMs classify by using the model given by
$h_{w,b}(x) = \textrm{sign}(\langle w, \phi(x)\rangle_{\mathcal{H}} + b)$

Different kernels can be used, e.g. the polynomial kernel or the 
RBF kernel $k(x,x') = e^{-\gamma || x-x' ||^2}$, where
$\gamma$ and $d$ are parameters of the kernel, that must
be tuned to the given data.

SMO.

On the practical side, LIBSVM can be regarded as the 
reference implementation of the SMO and the 'golden standard'
of SVM solvers \citep{cc01a}.

@@

There are only a few SVM libraries readily available from within \proglang{R}.
This includes the well-known packages like \pkg{kernlab} (via \code{ksvm()})~\citep{kernlab}, 
and \pkg{e1071} (via \code{svm()})~\citep{dimitriadou2008misc}
as well as other less known ones, e.g. \pkg{SwarmSVM}~\citep{SwarmSVM} 
and \pkg{lasvmR}~\citep{lasvmR}.
All of these link existing \proglang{C++} sources directly from within \proglang{R} and need
more or less extensive wrappers to allow the user direct access to
the options of the underlying SVM solver.

@@

\subsection{Wrapper}

A wrapper is a S3 object: It must provide routines
for assembling the command line call for training as well as testing,
reading and writing the model and a few general routines like searching
the binaries and printing.

Note that the search for binaries might have a huge toll on HPC clusters,
so it is adviced to specify the paths directly instead of relying on the 
automatism.

We provided default wrappers for LIBSVM, LASVM, CVM/BVM,
SVMperf, BSGD and LLSVM.

\section{Using a Wrapper}

To use the SVMBridge, one usually perform the following steps:
Add the external wrapper to the SVMBridge, load the data, train a model, use the
model for prediction.

Notice that the data must be written by the SVMBridge before calling 
the corresponding SVM Solver, as there is no way to pass data via memory
to command line tools. Therefore

\subsection{Adding the Wrapper to the SVMBridge}

The first step is to make the SVMBridge aware of a new wrapper.
This is simply done by calling \code{addSVMPackage()}.
There are several options.
The SVMBridge comes with an easy mechanism to search 
the corresponding binaries (specified in the wrapper) to ease
the usage. To use this, one can instead call \code{findSVMSoftware}.

\subsection{Reading Sparse Data}

Most SVM solver work with the sparse data format.
This format consists of:
Each line (delimited by a CR/LF) is given by a label
and the non-zero components of the data point,
e.g. to encode the vector six dimensional vector 
 $ ( 0 , 0 , 0 , 0.2 , 0 , 1.4 ) $
belonging to class $3$, the sparse 
data format would contain the line {3 4:0.2 6:1.4}.

Although reading these files into \proglang{R} is possible via either
the \pkg{kernlab} or the \pkg{e1071} routines, both only provide \proglang{R} solutions,
and therefore suffer from suboptimal performance.
The SVMBridge package provides a simple sparse data format 
reading and writing, which are implemented in \proglang{C++} and
therefore are nearly two orders of magnitudes faster than
the corresponding e1071 routines.
Notice, that currently only dense matrices are supported,
TODO: work around this with the Sparse Matrix Package.

\subsection{Training a model}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> model = trainSVM (method = "LIBSVM", cost = 1.0, gamma = 2.0,
			trainDataFile = "./data.sparse",
			epsilon = 0.042,
			modelFile = modelFile,
			useBias = TRUE,
			verbose = verbose
		)
\end{CodeInput}
\begin{CodeOutput}
...
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Predicions}

Predicting from a trained model is rather easy, as there are usually not many options.
As with training, the predict routine will accept a model either in memory or from a file.
The same is true for the data to predict.  Again notice that data and models in memory 
must be written to disk prior to calling the prediction binary.

\begin{CodeChunk}
\begin{CodeInput}
R> predictions = testSVM (model, data = '')
R> head(predictions)
\end{CodeInput}
\begin{CodeOutput}
...
\end{CodeOutput}
\end{CodeChunk}

\subsection{Optimization Values}

At times, some of the training values are of interest,
e.g. when comparing different solvers, the primal value
of the SVM problem as well as the dual value might  be of interest.
These can be computed via the \code{optimizationValues()} routine.

Specifically, this routine compute the following values:
\code{weight} is given by $\frac{1}{2} w^T w$, with
$w = \sum_i \alpha_i s_i$. Furthermore, we have
that $C \sum_i max (0, 1 - ..) $, and 
\code{primal} is then given by $primal = w + CHinge$.
\code{dual} on the other hand is simply
$w - \sum_i |\alpha_i|$. In general it must hold
that $primal \geq dual$. 
WHAT IF NOT? what does it mean?

\subsection{Helper  Routines}

A few routines have been placed into the package that were helpful
with testing. Namely, there are several generators for synthetical data sets,
collected from different publications. 
WHAT ELSE? can routines be used outside? detectTilde e.g.?

\subsection{Platform Considerations}

Although the SVMBridge was meant to be cross-platform, this goal
is hard to achieve. From a users perspective, it can be used on all three
major platforms (Linux, MacOS, Windows). Additional tests should
be started to make sure the package works as intended.

\subsection{Performance Considerations}

@@

\subsection{Training Call}

A simple S3 method has to be created that needs to be called
\code{createTrainingArguments.LIBSVM}. Apart from the very
first argument (taking the object itself), all other paramters
can be chosen freely. After training, from the output several
training information will be extracted by the 
\code{extractTrainingInfo()} function.

\subsection{Testing Call}

Similar to the training call, \code{createTestArguments()} needs to
writen. Again, it should accept three variables called testDataFile and
modelFile as well as predictionFile. There is also a
\code{extractTestInfo()} that will extract the important informations
from the output of the test binary.

\subsection{Handling Models}

As only LIBSVM models are being used, one can readily use the 
inbuild functions of the SVMBridge. If the solver works with its
own model, one needs to rewrite the readModel method of the wrapper.
This will only obtain the modelFile and will return the model
in the model format (S3 Object with XY, describe it somewhere else)
SVMBridge needs also a writeModel functions, as models passed in
memory needs to be written to disk. .
Finally it is necessary to access the predictions itself. For this one
needs to write the \code{readPredictions()} function. For LIBSVM this
will read the prediction files that is being written by LIBSVM to disk.

\subsection{Other functions}

Every wrapper should also be able to find itself in a given directory tree.
For this the S3 method \code{findSoftware()} must be provided.

At last, for querying reasons, there is also a print routine
so that wrapper can print out any further information about itself.
To see the currently loaded wrapper, one can call 
\code{overviewAllSVMSolver()}.

@@

We provided a simple framework, \pkg{SVMBridge},  to attach many SVM solvers to the \proglang{R} landscape.
\pkg{SVMBridge} does this by an easily extensible wrappers that manage the exchange of data and models
via command line.
By providing such a wrapper, linking any SVM solver to \proglang{R} becomes much easier.

>>>>

Kernelized Support Vector Machines (SVM) \cite{cortes1995support}
are binary classifiers that use a kernel $k$ to allow for non-linear
classification decisions
$x \mapsto \sign( \langle w, \phi(x) \rangle + b)$, where $\phi$ is the
(non-linear) feature map corresponding to the kernel
$k(x, x') = \langle \phi(x), \phi(x') \rangle$ in the reproducing kernel
Hilbert space $\mathcal{H}$.
We use the RBF kernel $k(x,x') = e^{-\gamma || x-x' ||^2}$, since it
yields excellent performance and enjoys a universal approximation
property.
The SVM training problem is given by
\begin{equation}
\min_{w \in \mathcal{H}, b \in \R} \quad \frac{1}{2} ||w||^2 + C \cdot \sum_{i=1}^n \max \Big( 0, 1 - y_i \big( \langle w, \varphi(x_i) \rangle_{\mathcal{H}} + b \big) \Big),
\label{primalproblem}
\end{equation}
where $\big\{(x_1, y_1), \dots, (x_n, y_n)\big\}$ are the labeled
training points and $C > 0$ is a regularization parameter that controls
the complexity of the predictive model.

The problem, although convex, can be solved by different means,
e.g. LIBSVM employs a dual
decomposition method, while  BSGD is a primal stochastic
gradient descent solver.

@@

In general, nearly all SVM solvers come as binary packages,
often written in C. Therefore it is not possible to use them 
directly from within \proglang{R}. Only a few SVM libraries are readily available,
includ LIBSVM inside the well-known packages like \pkg{kernlab} (via \code{ksvm()})~\citep{kernlab}, 
and \pkg{e1071} (via \code{svm()})~\citep{dimitriadou2008misc}
as well as other less known ones, e.g. \pkg{SwarmSVM}~\citep{SwarmSVM} 
and \pkg{lasvmR}~\citep{lasvmR}.
All of these link existing \proglang{C++} sources directly from within \proglang{R} and need
more or less extensive wrappers to allow the user direct access to
the options of the underlying SVM solver.

@@

\section{General worflow}

To use the SVMBridge, one usually perform the following steps:
Make the external wrapper incl. software binaries known to the SVMBridge, 
load the data, train a model, use the model for prediction.

\subsection{Wrappers}

A wrapper is a S3 object: It must provide routines
for assembling the command line call for training as well as testing,
reading and writing the model and a few general routines like searching
the binaries and printing.
We provided default wrappers for LIBSVM, LASVM, CVM/BVM,
SVMperf, BSGD and LLSVM.

Make the SVMBridge aware of a new wrapper is
simply done by calling \code{addSVMPackage()}.
The SVMBridge comes with an easy mechanism to search 
the corresponding binaries (specified in the wrapper) to ease
the usage. To use this, one can instead call \code{findSVMSoftware}.

\subsection{Data}

Most SVM solver work with the LIBSVM sparse data format.
This format saves the labels together with all non-zero indicies of every data point, e.g.
$(0, 0, 0, 0.2 , 0, 1.4)^T$  belonging to class $3$, will be encoded as \texttt{'3 4:0.2 6:1.4'}.

Although reading these files into \proglang{R} is possible via e.g. the 
\pkg{e1071} package, most only provide pure \proglang{R} solutions,
and therefore suffer from suboptimal performance.
The SVMBridge package provides a simple sparse data format 
reading and writing, which are implemented in \proglang{C++} and
therefore can be nearly two orders of magnitudes faster than
the corresponding e1071 routines.

\subsection{Models}

Internally models contain support vectors and their coefficients.

\subsection{Training an SVM model}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> model = trainSVM (method = "LIBSVM", cost = 1.0, gamma = 2.0,
	trainDataFile = "./data.sparse", epsilon = 0.042, modelFile = modelFile)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Testing a model}

%Predicting from a trained model is rather easy, as there are usually not many options.
Testing (or doing predictions) needs a trained SVM model as well as test data to work on.
As with training, testSVM will accept data and models either from memory or from a file.
%The same is true for the data to predict.  Again notice that data and models in memory 
%must be written to disk prior to calling the prediction binary.

\begin{CodeChunk}
\begin{CodeInput}
R> testObj = testSVM (model, testDataFile = './australian_scale', 
	readPredictions = TRUE)
\end{CodeInput}
\end{CodeChunk}

The returned object will contain meta information like testing times as well as the predictions.

\subsection{Other Considerations}

Although the SVMBridge was meant to be cross-platform, this goal
is hard to achieve. From a users perspective, it can be used on all three
major platforms (Linux, MacOS, Windows). The tests should
be executed to make sure the package works as intended.

@@

@@

We provided a simple framework, \pkg{SVMBridge}, to attach binary SVM solvers easily to the \proglang{R} landscape.
The \pkg{SVMBridge} does this by an easily extensible wrappers that manage the exchange of data and models
via command line.
%By providing such a wrapper, linking any SVM solver to \proglang{R} becomes much easier.
